Identifying the Tissue of Origin of Extracellular Vesicles Using RNA Expression Signatures

### 2015-09-16 poster/poster.Rnw
clarify figure legends and make other changes Monica suggested

\begin{document}
\acrodef{EV}[EV]{Extracellular Vesicle}
\acrodef{SVM}{Support Vector Machine}
\acrodef{ROC}{Receiver Operating Characteristic}
\acrodef{KNN}{K Nearest Neighbors}

@@

    we we trained a multiclass \ac{SVM} to accurately identify the
    tissue of origin, and present the results of the identification of
    reads from uterine and placental origin here. Tables showing
    tissue specific genes from the other %
    %
    \Sexpr{length(unique(na.omit(tissue.specific.genes[,tissue])))} %
    % 
    tissues can be found on \url{http://dla2.us/p/em2015}.

@@

  \block{Methods}{ %
    \Sexpr{star.logs[grepl("Number of input reads",field),sum(as.numeric(value))]} %
    %
    RNAseq reads from %
    % 
    \Sexpr{nrow(genes.training)+nrow(genes.testing)} %
    % 
    samples corresponding to %
    \Sexpr{length(unique(na.omit(tissue.specific.genes[,tissue])))} %
    different tissues obtained from the Roadmap Epigenomics
    \cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping} and GTEx
    \cite{Gtexconsortium.2015:Thegenotypetissueexpression} projects,
    as well as uterine (GSE50599,
    \cite{Chan.ea2014:Assessmentofmyometrialtranscriptome}) and
    placenta (GSE66622,
    \cite{Hughes.ea2015:Evaluatingintra-andinter-individual}) samples
    from the SRA.

@@

    The \ac{SVM} was generated using caret
    \cite{Kuhn2008:Buildingpredictivemodelsinr} and svmRadial
    \cite{Karatzoglou2004:kernlab} with 10 cycles of 10 repeats and a
    tune length of 8 with a training set representing $\approx 75\%$ of the
    samples ($\approx 2450$ samples total, including $\approx 35$
    uterine or $\approx 60$ placenta) in each of the groups chosen at
    random, with the remaining samples making up a testing set
    ($\approx 820$ samples). For comparison purposes, we also
    generated a \ac{KNN} based predictor using the same training and
    testing sets of genes.

@@

<<uterus.specific.markers.figure,fig.width=8,fig.height=8,echo=FALSE>>=
print(specific.markers.plot("uterus")
      + xlab("Uterus Specific Genes")
      + ylab("Tissue"))
@
}
\captionof{figure}{Uterus specific genes. $\log_2 \mathrm{FPKM}$ of
  genes with tissue specific indices $\ge 0.98$ which are specific to
  uterus tissue in uterus and tissues with second highest expression
  for those genes. With the exception of FTX\_1, which is expressed in
  multiple fetal tissues and MYH11, a major component of smooth
  muscle, all genes are highly specific to uterus.}

@@

}
\captionof{figure}{Placenta specific genes. $\log_2 \mathrm{FPKM}$ of
  genes with tissue specific indices $\ge 0.98$ and
  $\mathrm{FPKM} \ge 1024$ which are specific to uterus tissue in
  placenta and tissues with second highest expression for those
  genes.}
}

@@

<<placenta.roc.curve,fig.width=8,fig.height=5,echo=FALSE>>=
layout(matrix(1:2,1,2))
pl.roc.obj1 <- 
    plot.roc(as.numeric(predictions.caret.svm[["placenta"]][["testing"]][,"actual"])-1,
             as.numeric(predictions.caret.svm[["placenta"]][["testing"]][,"placenta"]),
             percent=TRUE,
             ## main="Training vs Testing set for Uterus",
             col="#002058"
             )
pl.roc.obj2 <- 
    lines.roc(as.numeric(predictions.caret.svm[["placenta"]][["training"]][,"actual"])-1,
              as.numeric(predictions.caret.svm[["placenta"]][["training"]][,"placenta"]),
              percent=TRUE,
              col="#E87722")
# testobj <- roc.test(rocobj1, rocobj2)
# text(50, 50, labels=paste("p-value =", format.pval(testobj$p.value)), adj=c(0, .5))

@@

\block{Conclusions}{
  \begin{itemize}
  \item Pregnancy-specific glycoproteins (PSG1-8) are specific to
    placental tissues,and PAEP Progestagen-Associated Endometrial
    Protein is specific to uterus, as previously described.
  \item Numerous Long-noncoding RNA and smaller RNAs are specific to uterus and placenta
  \item \ac{SVM} are able to accurately identify the tissue of origin
    for uterus and placenta.
  \item Very few sequenced reads are required to accurately identify
    tissues of Placental origin. We are able to detect 100\% of the
    placenta tissues with 80\% specificity with just 10 reads.
  \item Non-maternal variants may also be able to distinguish between
    placental reads and maternal reads in addition to the approach
    used here.
  \end{itemize}
}

>>>>

\begin{document}
\acrodef{EV}[EV]{Extracellular Vesicle}
\acrodef{SVM}{Support Vector Machine}
\acrodef{ROC}{Receiver Operating Characteristic}
\acrodef{KNN}{K Nearest Neighbors}
\acrodef{SRA}{Sequence Read Archive}

@@

    we we trained a multiclass \ac{SVM} to accurately identify the
    tissue of origin, and present the results of the identification of
    reads from uterine and placental origin here. Tables showing
    tissue specific genes from all %
    %
    \Sexpr{length(unique(na.omit(tissue.specific.genes[,tissue])))} %
    % 
    tissues can be found on \url{http://dla2.us/p/em2015}.

@@

  \block{Methods}{ %
    \Sexpr{star.logs[grepl("Number of input reads",field),sum(as.numeric(value))]} %
    %
    RNAseq reads from %
    % 
    \Sexpr{length(star.logs[grepl("Number of input reads",field),value])} %
    % 
    samples (supplemented with FPKMs from
    \Sexpr{nrow(predictions.caret.svm[["uterus"]][["training"]])+
      nrow(predictions.caret.svm[["uterus"]][["testing"]]) -
      length(star.logs[grepl("Number of input reads",field),value])} %
    %
    additional samples without raw data in the \ac{SRA})
    % 
    corresponding to %
    \Sexpr{length(unique(na.omit(tissue.specific.genes[,tissue])))} %
    different tissues were obtained from the Roadmap Epigenomics
    \cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping} and GTEx
    \cite{Gtexconsortium.2015:Thegenotypetissueexpression} projects,
    as well as uterine (GSE50599,
    \cite{Chan.ea2014:Assessmentofmyometrialtranscriptome}) and
    placenta (GSE66622,
    \cite{Hughes.ea2015:Evaluatingintra-andinter-individual}) samples
    from the \ac{SRA}.

@@

    The \ac{SVM} was generated using caret
    \cite{Kuhn2008:Buildingpredictivemodelsinr} and svmRadial
    \cite{Karatzoglou2004:kernlab} with 10 cycles of 10 repeats and a
    tune length of 8 with a training set representing $\approx 75\%$ of the
    samples ($\approx 2450$ samples total, including $\approx 35$
    uterine or $\approx 60$ placenta) in each of the groups chosen at
    random, with the remaining samples making up a testing set
    ($\approx 820$ samples).

@@

<<uterus.specific.markers.figure,fig.width=8,fig.height=8,echo=FALSE>>=
print(specific.markers.plot("uterus")
      + xlab("Uterus Specific Genes")
      + ylab("Tissue"))
@
}
\captionof{figure}{Expression of uterus specific genes in uterus and
  tissues with the second highest expression for those genes.
  Expression is $\log_2 \mathrm{FPKM}$ and genes are tissue specific
  if they have a tissue specific expression $\tau_g \ge 0.98$. The
  known uterus specific gene PAEP and short RNAs ZNFX1-AS1\_3 (87bp)
  and AL356310.1 (94bp) are highly uterus specific.}

@@

}
\captionof{figure}{Expression of placenta specific genes in placenta
  and tissues with the second highest expression for those genes.
  Expression is $\log_2 \mathrm{FPKM}$ and genes are tissue specific
  if they have a tissue specific expression $\tau_g \ge 0.98$ and
  $\mathrm{FPKM} \ge 1024$ in placenta. The known placenta specific
  PSG genes and CSH genes are all highly placenta specific, and short
  RNAs mir-126, mir-622, and PAPPA-AS1 are fairly placenta specific.}
}

@@

<<placenta.roc.curve,fig.width=8,fig.height=4,echo=FALSE>>=
layout(matrix(1:2,1,2))
pl.roc.obj1 <- 
    plot.roc(as.numeric(predictions.caret.svm[["placenta"]][["testing"]][,"actual"])-1,
             as.numeric(predictions.caret.svm[["placenta"]][["testing"]][,"placenta"]),
             percent=TRUE,
             ## main="Training vs Testing set for Uterus",
             col="#002058"
             )
pl.roc.obj2 <- 
    lines.roc(as.numeric(predictions.caret.svm[["placenta"]][["training"]][,"actual"])-1,
              as.numeric(predictions.caret.svm[["placenta"]][["training"]][,"placenta"]),
              percent=TRUE,
              col="#E87722")
# testobj <- roc.test(rocobj1, rocobj2)
# text(50, 50, labels=paste("p-value =", format.pval(testobj$p.value)), adj=c(0, .5))

@@

\block{Conclusions}{
  \begin{itemize}
  \item Pregnancy-specific glycoproteins (PSG1-8) are specific to
    placental tissues, and PAEP is specific to uterus, as previously
    described.
%   \item Numerous Long-noncoding RNA and smaller RNAs are specific to uterus and placenta
  \item \ac{SVM} are able to accurately identify the tissue of origin
    for uterus and placenta.
  \item Very few sequenced reads are required to accurately identify
    tissues of Placental origin. We are able to detect 100\% of the
    placenta tissues with 80\% specificity with just 10 reads.
  \item We can easily identify exosomes originating from the placenta,
    as even the smallest exosomes can contain 70 RNA molecules
    \cite{Li.ea2014:AnalysisofRNAcontent}.
  \item Non-maternal variants may also be able to distinguish between
    placental reads and maternal reads.
  \end{itemize}
}
